Caustic: Reinventing Database Transactions

### 2017-09-24 caustic-assets/paper/main.tex
Removed IEEE stylesheet
Syntax Update
Initial implementation of the Caustic compiler. Still deeply flawed, but
the general outline of it has been written.

\documentclass[journal]{../styles/IEEEtran}

@@

% Abstract.
\begin{abstract}
Database languages are designed for ease of implementation and not ease of use. For example, every relational database has its own interpretation of the SQL specification that is most convenient for the underlying database architecture and most often incompatible with its competition, and most key-value stores expose primtiive interfaces that are lacking in functionality and programmability. Caustic is a robust query language that is intuitive to use and straightforward to integrate with any database.
\end{abstract}

% Introduction
\section{Introduction}
The syntax of query languages varies wildly between databases. For example, relational databases like MySQL and PostgreSQL claim to implement the same SQL specification, but, in reality, all but the most trivial of queries are incompatible with each other. This lack of uniformity between languages tightly couples the database with the queries that are run on it, and makes it difficult to switch storage systems as requirements change.

The structure of query languages is deeply motivated by the capabilities and limits of the underlying database and not by what applications actually require. In particular, the language of relational algebra, which SQL is based upon, is not first-nature for most programmers. [On the Principles of Good Language Design, CJ Date] [http://wiki.c2.com/?SqlFlaws]

It's clear that query languages are typically designed for ease of implementation and not ease of use. They have disparate interfaces, deficient functionality, and unintuitive syntax. Caustic provides both a robust language for specifying transactions and an efficient runtime for executing them on arbitrary databases.

% Background
\section{Background}

% Runtime
\section{Runtime}
The runtime is responsible for executing transactions on arbitrary databases. Transactions are represented as abstract-syntax trees, because they are easy to manipulate, serialize, and optimize. The runtime exposes a Thrift interface for cross-language transaction execution, and is capable of running on any database that supports the following primitive operations:

\begin{itemize}
  \item \textbf{get}: Returns the values of a set of keys.
  \item \textbf{cput}: Conditionally apply changes iff dependent keys remain unchanged.
\end{itemize}

\subsection{Representation}

\subsection{Snapshot Isolation}

\subsection{Optimizations}
Because transactions are represented as abstract-syntax trees, the runtime is able to perform some interesting optimizations to improve execution performance. Most importantly, the runtime performs read and write batching. In other words, the runtime fetches \emph{as many reads as possible} and applies \emph{all writes together}. Clearly, the runtime performs optimally few roundtrips to and from the underlying key-value store, which is typically the performance bottleneck in database applications.

\subsection{Tunable Consistency}
Transactions are executed using MVCC, and derive their consistency and isolation guarantees from the underlying key-value store. If the underlying key-value has a repeatable isolation level, for example, then so will transactions in Caustic. If the underlying key-value is not transactional, then neither are transactions in Caustic. Therefore, applications may substitute a key-value store with weaker transactional guarantees (or no transactions at all!) to improve performance.

\subsection{Benchmarks}
Compare the Caustic runtime (running on RocksDB) to comparable systems like Calvin and Tango. Rely on quantitative metrics like end-to-end latency and transaction throughput on standard database benchmarks like YCSB and TPC-C.

\subsubsection{Effect of Transaction Size}
Performance scales linearly with transaction size. This is enabled by the tail recursive design of the runtime and can be verified empirically from benchmarks.

\subsubsection{Effect of Read or Write Skew}
Transactions with more reads than writes will have a different performance profiles than transactions with more writes than reads. However, this difference in performance is an artifact of the choice of underlying key-value store and not due to a read or write bias in the runtime. In other words, a read optimized key-value store will fare better with read-heavy transactions and a write optimized key-value store with write-heavy transactions in Caustic.

@@

% Syntax
\section{Syntax}
The Caustic syntax is a new high-level programming language that compiles into transactions on the Caustic runtime. Regardless of the features they might have (objects, functions, types, etc.), all programs eventually compile down into load, stores, and operations on memory - and memory is a key-value store. The purpose of Caustic syntax is to provide a rich language for expressing \emph{transactional} loads, stores, and operations on \emph{key-value stores}.

\subsection{Concurrency}
Because all programs written in Caustic compile into transactions, all programs are automatically thread-safe. No fancy synchronization mechanisms are required to build parallelizable programs in Caustic, everything is concurrent by default.

\subsection{Separation of Code and Data}
Every database has its own language for writing stored procedures. Even SQL databases typically support very different subsets of the SQL specification, that are not interoperable for anything but the most trivial of programs.  Because the syntax varies so wildly between different databases, stored procedures are very tightly coupled with the choice of underlying database. This makes it tremendously expensive to change databases once development is under way, and impossible to test stored procedures in isolation. Caustic provides a uniform interface over arbitrary databases. In other words, Caustic programs written for ione database will be identical to the same program written for any another. This allows programmers to select whichever database is most appropriate for their use-case (eg. in-memory for tests, MySQL for production).

@@

\subsection{Benchmarks}
Compare the Caustic syntax qualitatively to SQL, Tango, and Java. Measure differences in code complexity (measured in Cyclomatic or Halstead Complexity), size of executables, and lines of code for a variety of sample use cases (eg. implementing ZooKeeper or BookKeeper).

>>>>

\documentclass[12pt]{article}

@@

% Abstract.
\begin{abstract}
Databases are either easy to use or to scale. For example, relational databases were popularized in
large part because of the programmability of SQL. However, this programmability comes at an
incredible expense. Relational databases are notoriously difficult to support at scale, and so
developers have increasingly turned toward more specialized NoSQL systems that scale well be
shedding functionality. Developers are not only forced to choose between productivity and
performance, but also stark differences between query languages makes their choice of database
effectively permanent. Even databases that claim to support the same SQL standard only implement
incompatible subsets of its functionality. The lack of a truly uniform interface tightly couples
databases and the procedures that are executed against them. Caustic is both a syntax for expressing
transactions and a runtime for executing them on arbitrary key-value stores, that is both
straightforward to use and simple to integrate.
\end{abstract}

% Runtime
\section{Runtime}
  The runtime is responsible for executing transactions on arbitrary key-value stores. Runtimes
  implement Multiversion Concurrency Controls over any key-value store that satisfies an minimal
  interface.

  \subsection{Representation}
    A \texttt{Transaction} is represented by an abstract-syntax tree. A \texttt{Literal} corresponds
    to a leaf of this syntax tree, and an \texttt{Expression} corresponds to a node. The runtime
    respects three different types of literals (\texttt{Real}, \texttt{Text}, and \texttt{Flag}) and
    thirty different kinds of expressions. With just these types and operations, the runtime is able
    to emulate a modern programming language complete with conditional branching, loops, and local
    variables.

  \subsection{Execution}
    Transactions are executed through repeated partial evaluation, according to the following
    procedure.

    \begin{itemize}
      \item \textbb{Fetch}: Call \texttt{get} on all keys that are read and written by the
        transaction and add the returned revisions to a local snapshot. In order to guarantee
        snapshot isolation, \texttt{get} is only called when a key is read or written for the
        first time. By batching reads and writes and avoiding duplicate fetches, the runtime is
        guaranteed to perform a minimal number of roundtrips to and from the database.
      \item \textbb{Evaluate}: Recursively replace all expressions that have literal operands with
        their corresponding literal result. For example, add(real(1), sub(real(0), real(2)))
        returns real(-1). The result of all write expressions is saved to a local buffer and all
        read expressions return the latest value of the key in either the local buffer or snapshot.
      \item \textbb{Repeat}: Re-evaluate the transaction until it reduces to a single literal
        value. Because all expressions with literal operands return a literal value, all
        transactions eventually reduce to a literal.
    \end{itemize}

  \subsection{Integrated Caching}
    MVCC allows the runtime to cache incoherently without sacrificing data integrity. The runtime
    speculates about the value of a key by reading a potentially stale version from cache, and then
    validate the cached versions when they commit. Multilevel, write-through caching is directly
    integrated into the runtime:

    \begin{itemize}
      \item Keys are updated on cache miss.
      \item Keys are automatically updated after successful transactions.
      \item Keys are automatically invalidated after conflicting transactions.
    \end{itemize}

    \subsubsection{Effect of Contention}
      While the implementation of caching is fault tolerant, it is particularly vulnerable to
      thrashing under contention. Consider if runtime A and runtime B read a key K from cache. A
      updates the value of K, and then subsequently updates the value of K in cache. When B tries to
      update the value of K it will cause a conflict, and so it will subsequently invalidate K in
      cache.

      A potential solution might be to remove the invalidation step from the procedure; however, the
      implementation would no longer be fault tolerant. Consider if A updates K but fails to write
      the new value to cache. Now the value in cache will remain stale until its evicted, but a key
      under high contention may never be evicted from cache.

  \subsection{Performance}

    \subsubsection{Effect of Transaction Size}
      The runtime has been empirically shown to scale linearly with transaction size.

    \subsubsection{Effect of Read or Write Skew}
      The quantity of read and write expressions will surely impact the performance profile of the
      runtime. However, the difference in performance is an artifact of the choice of underlying
      key-value store and not due to any inherent bias in the runtime.

@@

% Syntax
\section{Syntax}
  While the language exposed by the runtime is robust and fully-featured, it can become verbose and
  unintuitive as transactions get larger in size and broader in scope. In this section, we'll
  introduce the Caustic programming language.

  \subsection{Why Caustic?}
    Relational databases have remained dominant particularly because of the programmability of SQL.
    SQL provides developers with robust schemas that make it easy to structure and debug database
    queries. However, relational database are notoriously difficult to support at scale, and so
    developers are increasingly turning towards more specialized NoSQL systems that scale well by
    shedding functionality.

    While SQL has previously enjoyed remarkable popularity, the language is deeply flawed. In his
    letter entitled "Some Principles of Good Language Design", CJ Date outlines a number of these
    flaws as well as a criteria for evaluating languages in general. Date argues that well-designed
    languages should have a consistent syntax, a canonical implementation, a terse grammar, and an
    intuitive structure and provides examples that show how SQL fails to satisfy each of these
    fundamental requirements.

  \subsection{Automatic Concurrency}
    Because all programs written in Caustic compile into transactions on the Caustic runtime, all
    programs are automatically thread-safe. Developers no longer have to worry about fancy
    synchronization mechanisms in order to build parallelizable programs.

  \subsection{Uniform Interface}
    Every database has its own query language. Even databases that implement the same SQL
    specification usually support disparate subsets of the language. This has the effect of tightly
    coupling the underlying database and the procedures that are run against it. Caustic provides
    a uniform interface over arbitrary databases, which allows programmers to select whichever
    underlying database is most appropriate for their use-case without modifying the code in any
    way shape or form.

@@
